Design and Implementation of Probabilistic Programming Language Anglican

### 2016-11-27 paper/paper.tex
minor: indentation in code block off by 1 space
update

\title{Design and Implementation of Probabilistic Programming Language Anglican}
\authorinfo{David Tolpin \and Jan-Willem van de Meent \and Hongseok Yang \and Frank Wood}{University of Oxford}{dtolpin@robots.ox.ac.uk \and jwvdm@robots.ox.ac.uk \and hongseok.yang@cs.ox.ac.uk \and fwood@robots.ox.ac.uk}

@@

On top of these two approaches are methods in which a new
language is implemented `inside' another language of the same
level of abstraction. Different languages provide different
means for this; Lisp is famous for the macro facility
that allows to extend the language almost without
restriction --- by writing \textit{macros}, one adds new
constructs to the existing language. There are several uses of
macros --- one is to extend the language \textit{syntax}, for
example, by adding new control structures; another is to keep
the existing syntax but alter the \textit{operational semantics}
--- the way programs are executed and compute their outputs.

@@

Literals are the same in Anglican and Clojure. They are
left unmodified; literals are a subset of opaque expressions.
However, the Clojure syntax has a peculiarity
of using the syntax of compound literals (vectors, hash maps,
and sets) for data constructors. Hence, compound literals must
be traversed recursively, and if there is a nested non-literal
component, transformed into a call to the corresponding data
constructor. Functions \texttt{cps-of-vector},
\texttt{cps-of-hash-map}, \texttt{cps-of-set}, called from
\texttt{cps-of-expression}, transform Clojure constructor syntax
(\texttt{[...]}, \texttt{\{...\}}, \texttt{\#\{...\}}) into the
corresponding calls:
\begin{lstlisting}[style=default]
=> (cps-of-vector [0 1 2] 'cont)
(cont (vector 0 1 2) $state)
=> (cps-of-hash-map {:a 1 :b 2} 'cont)
(cont (hash-map :a 1, :b 2) $state)
=> (cps-of-set #{0 1} 'cont)
(cont (set (list 0 1)) $state)
\end{lstlisting}

@@

Memoization is often implemented on top of a mutable
dictionary, where the key is the argument list and the value is
the returned value. However, there are no mutable data
structures in a probabilistic program, hence \texttt{mem}'s
memory is stored as a nested dictionary in the program state
introduced during CPS transformation (function \texttt{mem-cps}).  
Every memoized function gets a
unique automatically generated identifier. Each time a memoized
function is called, one of two continuations is chosen,
depending on whether the same function (a function with the same
identifier) was previously called in the same run of the
probabilistic program with the same arguments. If the memoized
result is available, the continuation of the memoized function
call is immediately called with the stored result. Otherwise,
the argument of \texttt{mem} is called with a continuation which
first creates an updated state with the memoized result, and
then calls the `outer' continuation with the result and the
updated state:
\begin{lstlisting}[style=default]
=> (mem-cps '(foo))
(let [M (gensym "M")]
  (fn mem23623 [C $state & P]
    (if (in-mem? $state M P)
      ;; previously memoized result
      (fn []
        (C (get-mem $state M P) $state))
      ;; new computation
      (clojure.core/apply
        foo
        ;; memoize result in state
        (fn [V $state]
          (fn [] (C V (set-mem $state M P V))))
        $state
        P)))))
\end{lstlisting}

@@

The sole purpose of the algorithm identifier of \texttt{infer} is to invoke
an appropriate overloading or implementation of the function --- conventionally,
the identifier is a Clojure keyword (a symbolic constant starting with colon)
related to the algorithm name, such as \texttt{:lmh} for Lightweight
Metropolis-Hasting and \texttt{:pcascade} for Particle Cascade.
The second parameter is a query as defined by
the \texttt{defquery} form or its anonymous version \texttt{query}. For instance,
the following Clojure code invokes \texttt{infer} on an Anglican query
defined anonymously via the \texttt{query} form:
%If needed, the query
%can be defined anonymously, right in the argument list of a call
%to \texttt{infer}:
\begin{lstlisting}[style=default]
(let [x 1]
  (infer :pgibbs (query x) nil))
\end{lstlisting}
A query is executed by calling the initial continuation of the
query, which accepts a value and a state. The state is supplied
by the inference algorithm, while the value is provided as a
parameter of \texttt{infer}. A query does not have to have any
parameters, in which case the value can be simply \texttt{nil}.
When a query is defined with a binding for the initial value,
the value becomes available inside the query. A query may
accept multiple parameters using Clojure's structured binding.
For instance, it may take multiple parameters as components
of an input vector. In this case, the initial value is given
as a structured value, such as a vector, and the components
of this value are matched to the corresponding
parameters of the query via the destructuring mechanism of Clojure.
%The value is destructured via pattern match in case of a structured binding.
For example,
\begin{lstlisting}[style=default]
(defquery my-query [mean sd]
  (sample (normal mean sd)))

@@

Checkpoints can be uniquely named at compilation time;
however, at runtime a checkpoint corresponding to a single code
location may occur multiple time due to recurrent invocation of
the function containing the checkpoint. Every unique occurrence
of a checkpoint must receive a different address. An addressing
scheme based on computing of stack addresses of checkpoints was
described in the context of Lightweight
Metropolis-Hastings~\cite{WSG11}.  This scheme has advantages
over the naive scheme where dynamic occurrences of checkpoints
are numbered sequentially. However, it impacts the
performance of probabilistic programs because of the computation
cost of computing the stack addresses:
\begin{enumerate}
    \item On every function call, a component is added to the
        address, hence the address size is linear in stack depth.
    \item Every function call must be augmented by symbolic
        information required to compute stack addresses.
\end{enumerate}
In addition, the above scheme can still lead to inferior
correspondence between checkpoints in different traces: in
Anglican and other probabilistic languages where distributions
are first-class objects checkpoints with incompatible arguments
can correspond to the same stack address. Consider the following
program fragment:
\begin{lstlisting}[style=default]
(let [dist (if use-gamma
               (gamma 2. 2.)
               (normal 0. 1.))]
  (sample dist))
\end{lstlisting}
The \texttt{sample} checkpoint has the same stack address in
different traces. However, the random values should not be
reused between different distributions. Further on, in some
algorithms, such as black-box variational inference, the role of
checkpoint addresses is semantic rather than heuristic ---
appropriate correspondence must be established
between checkpoints in different traces for the algorithm to
work.

@@

Similar to distributions, \textit{random processes}, related to
so called `exchangeable random procedures'\footnote{However, random sequences
generated by Anglican random processes are not required to be
exchangeable.}, are defined using
macro \texttt{defproc} and implement the
\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol (that is,
an interface for these random-sequence objects). The protocol
has two methods --- \texttt{produce}, which returns a
distribution object, and \texttt{absorb}, which returns a new
random process object obtained by absorbing a value drawn from
the process. Here is a definition of the Poisson process:
\begin{lstlisting}[style=default]
(defproc PP
  "Poisson point process"
  [rate] [dist (exponential rate)]
  (produce [this] dist)
  (absorb [this value] this))
\end{lstlisting}
While distributions are random and non-functional,
random processes are deterministic and functional, hence
\texttt{produce} and \texttt{absorb} are called directly and do
not have corresponding special forms in Anglican:
\begin{itemize}
    \item Values produced by the process are sampled from the distribution
        returned by \texttt{produce}.
    \item When a value is observed from the distribution, a new
        process, with the observed value absorbed, is returned
        by \texttt{absorb}.
\end{itemize}
A common pattern of programming with processes in Anglican is
recursively computing a new process based on the observed value:
\begin{lstlisting}[style=default]
(loop [process (HawkesPointProcess)
       [value0 & values0] values] 
  (let [dist (produce process)]
    (observe dist value0)
    (recur (absorb process value0) values0)))
\end{lstlisting}
Unlike conventional implementations of random processes,
Anglican's random processes do not have mutable state.

@@

\begin{itemize}
    \item There is an adjacent office quarter, and it takes
        between 5 and 15 minutes on average to walk from an
        office to the deli, were different average times are  for
        different buildings in the quarter.
    \item Depending on traffic lights, the walking time varies by about 2
        minutes.
    \item The lunch break starts at 1:00pm, and the workday ends at 6:00pm.
    \item The waiter's odds that this is the same customer are 2 to 1.
\end{itemize}

@@

For inference, one often chooses known distribution to
represent uncertainty. We choose the normal
distribution for representing uncertainty about average arrival
time.

@@

There are two possibilities: either the same customer visited
the deli twice, or two different customers came to the deli,
one for lunch, one for dinner. We define an \textbf{Anglican} function
for each case. Note that this is the first time we switch from
Clojure to Anglican. The functions must be written in Anglican
(and hence defined using \texttt{defm} instead of \texttt{defn})
because they contain probabilistic forms \texttt{sample} and
\texttt{observe}.

@@

Both functions have the same structure: we first `guess' the
average arrival time and then observe the actual time from a
distribution parameterized by the guessed time. However, in
\texttt{same-customer} the average arrival time is the same for
both the lunch and the dinner, while for
\texttt{different-customers} two average arrival times are
guessed independently.

@@

\begin{lstlisting}[style=default]
(defquery deli [time-to-arrive-prior
                lunch-delay
                dinner-delay]
  (let [is-same-customer (sample (flip p-same))
        observe-customer (if is-same-customer
                            same-customer
                            different-customer)]
    {:is-same-customer is-same-customer
     :times-to-arrive (observe-customer 
                        time-to-arrive-prior
                        lunch-delay
                        dinner-delay)}))
\end{lstlisting}

@@

To approximate the inferred distribution, we extract a finite
subsequence of samples. Many algorithms use an initial
subsequence of samples to converge to the target distribution,
hence we drop initial $N$ samples ($N$ is 5000 in the code
snippet below), and collect the next $N$ samples.

@@

\begin{lstlisting}[style=default]
(def p-same+ 
     (/ (count (filter :same-customer results))
        (double N)))
\end{lstlisting}

With the specified observations, \texttt{-same+} is $\approx
0.12$. The results may vary from run to run, and, for given
algorithm, the accuracy depends on the number of samples we
chose to retrieve. The probability is much lower than the
waiter's guess \texttt{p-same} ($\frac 2 3$).

@@

\begin{lstlisting}[style=default]
;; single customer                      
(def time-to-arrive+ 
     (map (comp first :times-to-arrive)
          (filter :same-customer results)))
(def mean-to-arrive+ (mean time-to-arrive+))
(def sd-to-arrive+ (std time-to-arrive+))

;; two customers
(def times-to-arrive+ 
     (map :times-to-arrive 
          (filter (complement :same-customer)
                  results)))
(def mean-1-to-arrive+
     (mean (map first times-to-arrive+)))
(def sd-1-to-arrive+
     (std (map first times-to-arrive+)))
(def mean-2-to-arrive+
     (mean (map second times-to-arrive+)))
(def sd-2-to-arrive+
     (std (map second times-to-arrive+)))
\end{lstlisting}

In addition to plotting the distribution histograms, we use
functions \texttt{mean}, \texttt{std} provided, along with other
useful statistical functions, in the \texttt{anglican.{\linebreak[0]}stat}
namespace. This is another illustration of advantage of tight
integration between Clojure and Anglican --- probabilistic
models are expressed in Anglican. However, processing of data and
results can rely on the full power of Clojure.

>>>>

\title{Design and Implementation of Probabilistic Programming Language Anglican}
\authorinfo{David Tolpin \and Jan Willem van de Meent \and Hongseok Yang \and Frank Wood}{University of Oxford}{dtolpin@robots.ox.ac.uk \and jwvdm@robots.ox.ac.uk \and hongseok.yang@cs.ox.ac.uk \and fwood@robots.ox.ac.uk}

@@

On top of these two approaches are methods in which a new
language is implemented `inside' another language of the same
level of abstraction. Different languages provide different
means for this; Lisp is notorious for the macro facility
that allows to extend the language almost without
restriction --- by writing \textit{macros}, one adds new
constructs to the existing language. There are several uses of
macros --- one is to extend the language \textit{syntax}, for
example, by adding new control structures; another is to keep
the existing syntax but alter the \textit{operational semantics}
--- the way programs are executed and compute their outputs.

@@

Literals are the same in Anglican and Clojure. They are
left unmodified; literals are a subset of opaque expressions.
However, the Clojure syntax has a peculiarity
of using the syntax of compound literals (vectors, hash maps,
and sets) for data constructors. Hence, compound literals must
be traversed recursively, and if there is a nested non-literal
component, transformed into a call to the corresponding data
constructor. Functions \texttt{cps-of-vector},
\texttt{cps-of-hash-map}, \texttt{cps-of-set}, called from
\texttt{cps-of-expression}, transform Clojure constructor syntax
(\texttt{[...]}, \texttt{\{...\}}, \texttt{\#\{...\}}) into the
corresponding calls:
\begin{lstlisting}[style=default]
=> (cps-of-vector [0 1 2] 'cont)
(cont (vector 0 1 2) $state)
=> (cps-of-hash-map {:a 1, :b 2} 'cont)
(cont (hash-map :a 1, :b 2) $state)
=> (cps-of-set #{0 1} 'cont)
(cont (set (list 0 1)) $state)
\end{lstlisting}

@@

Memoization is often implemented on top of a mutable
dictionary, where the key is the argument list and the value is
the returned value. However, there are no mutable data
structures in a probabilistic program. Hence, \texttt{mem}'s
memory is stored as a nested dictionary in the program state
introduced during CPS transformation (function \texttt{mem-cps}).  
Every memoized function gets a
unique automatically generated identifier. Each time a memoized
function is called, one of two continuations is chosen,
depending on whether the same function (a function with the same
identifier) was previously called in the same run of the
probabilistic program with the same arguments. If the memoized
result is available, the continuation of the memoized function
call is immediately called with the stored result. Otherwise,
the argument of \texttt{mem} is called with a continuation which
first creates an updated state with the memoized result, and
then calls the `outer' continuation with the result and the
updated state:
\begin{lstlisting}[style=default]
=> (mem-cps '(foo))
(let [M (gensym "M")]
  (fn mem23623 [C $state & P]
    (if (in-mem? $state M P)
      ;; previously memoized result
      (fn []
        (C (get-mem $state M P) $state))
      ;; new computation
      (clojure.core/apply
        foo
        ;; memoize result in state
        (fn [V $state]
          (fn [] (C V (set-mem $state M P V))))
        $state
        P)))))
\end{lstlisting}

@@

The sole purpose of the algorithm identifier of \texttt{infer} is to invoke
an appropriate overloading or implementation of the function --- conventionally,
the identifier is a Clojure keyword (a symbolic constant starting with colon)
related to the algorithm name, such as \texttt{:lmh} for Lightweight
Metropolis-Hastings and \texttt{:pcascade} for Particle Cascade.
The second parameter is a query as defined by
the \texttt{defquery} form or its anonymous version \texttt{query}. For instance,
the following Clojure code invokes \texttt{infer} on an Anglican query
defined anonymously via the \texttt{query} form:
%If needed, the query
%can be defined anonymously, right in the argument list of a call
%to \texttt{infer}:
\begin{lstlisting}[style=default]
(let [x 1]
  (infer :pgibbs (query x) nil))
\end{lstlisting}
A query is executed by calling the initial continuation of the
query, which accepts a value and a state. The state is supplied
by the inference algorithm, while the value is provided as a
parameter of \texttt{infer}. A query does not have to have any
parameters, in which case the value can be simply \texttt{nil}.
When a query is defined with a binding for the initial value,
the value becomes available inside the query. A query may
accept multiple parameters using Clojure's structured binding.
For instance, it may take multiple parameters as components
of an input vector. In this case, the initial value is given
as a structured value, such as a vector, and the components
of this value are matched to the corresponding
parameters of the query via the destructuring mechanism of Clojure.
%The value is destructured via pattern match in case of a structured binding.
For example,
\begin{lstlisting}[style=default]
(defquery my-query [mean sd]
  (sample (normal mean sd)))

@@

Checkpoints can be uniquely named at compilation time;
however, at runtime a checkpoint corresponding to a single code
location may occur multiple time due to recurrent invocation of
the function containing the checkpoint. Every unique occurrence
of a checkpoint must receive a different address. An addressing
scheme based on computing of stack addresses of checkpoints was
described in the context of Lightweight
Metropolis-Hastings~\cite{WSG11}.  This scheme has advantages
over the naive scheme where dynamic occurrences of checkpoints
are numbered sequentially. However, it impacts the
performance of probabilistic programs because of the computation
cost of computing the stack addresses:
\begin{enumerate}
    \item On every function call, a component is added to the
        address. Hence, the address size is linear in stack depth.
    \item Every function call must be augmented by symbolic
        information required to compute stack addresses.
\end{enumerate}
In addition, the above scheme can still lead to inferior
correspondence between checkpoints in different traces: in
Anglican and other probabilistic languages where distributions
are first-class objects checkpoints with incompatible arguments
can correspond to the same stack address. Consider the following
program fragment:
\begin{lstlisting}[style=default]
(let [dist (if use-gamma
               (gamma 2. 2.)
               (normal 0. 1.))]
  (sample dist))
\end{lstlisting}
The \texttt{sample} checkpoint has the same stack address in
different traces. However, the random values should not be
reused between different distributions. Further on, in some
algorithms, such as black-box variational inference, the role of
checkpoint addresses is semantic rather than heuristic ---
appropriate correspondence must be established
between checkpoints in different traces for the algorithm to
work.

@@

Similar to distributions, \textit{random processes}, related to
so called `exchangeable random procedures'\footnote{However, random sequences
generated by Anglican random processes are not required to be
exchangeable.}, are defined using
macro \texttt{defproc} and implement the
\texttt{anglican.runtime/{\linebreak[0]}random-process} protocol (that is,
an interface for these random-sequence objects). The protocol
has two methods --- \texttt{produce}, which returns a
distribution object, and \texttt{absorb}, which returns a new
random process object obtained by absorbing a value drawn from
the process. Here is a definition of the Poisson process:
\begin{lstlisting}[style=default]
(defproc PP
  "Poisson point process"
  [rate] [dist (exponential rate)]
  (produce [this] dist)
  (absorb [this value] this))
\end{lstlisting}
While distributions are random and non-functional,
random processes are deterministic and functional. Hence,
\texttt{produce} and \texttt{absorb} are called directly and do
not have corresponding special forms in Anglican:
\begin{itemize}
    \item Values produced by the process are sampled from the distribution
        returned by \texttt{produce}.
    \item When a value is observed from the distribution, a new
        process, with the observed value absorbed, is returned
        by \texttt{absorb}.
\end{itemize}
A common pattern of programming with processes in Anglican is
recursively computing a new process based on the observed value:
\begin{lstlisting}[style=default]
(loop [process (HawkesPointProcess)
       [value0 & values0] values] 
  (let [dist (produce process)]
    (observe dist value0)
    (recur (absorb process value0) values0)))
\end{lstlisting}
Unlike conventional implementations of random processes,
Anglican's random processes do not have mutable state.

@@

\begin{itemize}
    \item There is an adjacent office quarter, and it takes
        between 5 and 15 minutes on average to walk from an
        office to the deli, where different average times are  for
        different buildings in the quarter.
    \item Depending on traffic lights, the walking time varies by about 2
        minutes.
    \item The lunch break starts at 1:00pm, and the workday ends at 6:00pm.
    \item The waiter's odds that this is the same customer are 2 to 1.
\end{itemize}

@@

For inference, one often chooses a known distribution to
represent uncertainty. We choose the normal
distribution for representing uncertainty about average arrival
time.

@@

There are two possibilities: either the same customer visited
the deli twice, or two different customers came to the deli,
one for lunch and the other for dinner. We define an \textbf{Anglican} function
for each case. Note that this is the first time we switch from
Clojure to Anglican. The functions must be written in Anglican
(and hence defined using \texttt{defm} instead of \texttt{defn})
because they contain probabilistic forms \texttt{sample} and
\texttt{observe}.

@@

Both functions have the same structure: we first `guess' the
average arrival time and then observe the actual time from a
distribution parameterized by the guessed time. However, in
\texttt{same-customer} the average arrival time is the same for
both the lunch and the dinner, while in
\texttt{different-customers} two average arrival times are
guessed independently.

@@

\begin{lstlisting}[style=default]
(defquery deli [time-to-arrive-prior
                lunch-delay
                dinner-delay]
  (let [is-same-customer (sample (flip p-same))
        observe-customer (if is-same-customer
                            same-customer
                            different-customers)]
    {:same-customer is-same-customer,
     :times-to-arrive (observe-customer 
                        time-to-arrive-prior
                        lunch-delay
                        dinner-delay)}))
\end{lstlisting}

@@

To approximate the inferred distribution, we extract a finite
subsequence of samples. Many algorithms use an initial
subsequence of samples to converge to the target distribution.
Hence, we drop initial $N$ samples ($N$ is 5000 in the code
snippet below), and collect the next $N$ samples.

@@

\begin{lstlisting}[style=default]
(def p-same+ 
     (/ (count (filter :same-customer results))
        (double N)))
\end{lstlisting}
The \texttt{:same-customer} here represents a function
that looks up the entry with the same name in a given map.
The map stores the result of a single sampled execution,
and the \texttt{:same-customer} entry in the map
records whether the same customer visits the deli in the execution.
The \texttt{filter} function in \texttt{p-same+} 
goes through all the maps in \texttt{results}
and picks only the ones whose executions involve just one customer.

With the specified observations, \texttt{p-same+} is $\approx
0.12$. The probability is much lower than the
waiter's guess \texttt{p-same} ($\frac 2 3$).
Of course, the results may vary from run to run, and, for a given
algorithm, the accuracy depends on the number of samples we
decide to retrieve.

@@

\begin{lstlisting}[style=default]
;; single customer                      
(def time-to-arrive+ 
    (map (fn [x] (first (:times-to-arrive x)))
         (filter :same-customer results)))
(def mean-to-arrive+ (mean time-to-arrive+))
(def sd-to-arrive+ (std time-to-arrive+))

;; two customers
(def times-to-arrive+ 
     (map :times-to-arrive 
          (filter  
            (fn [x] (not (:same-customer x)))
            results)))
(def mean-1-to-arrive+
     (mean (map first times-to-arrive+)))
(def sd-1-to-arrive+
     (std (map first times-to-arrive+)))
(def mean-2-to-arrive+
     (mean (map second times-to-arrive+)))
(def sd-2-to-arrive+
     (std (map second times-to-arrive+)))
\end{lstlisting}

In addition to plotting the distribution histograms, we use
functions \texttt{mean}, \texttt{std} provided
in the \texttt{anglican.{\linebreak[0]}stat} namespace
along with other
useful statistical functions.
This is another illustration of advantage of tight
integration between Clojure and Anglican --- probabilistic
models are expressed in Anglican. However, processing of data and
results can rely on the full power of Clojure.
